---
title: "Figure S — Single-cell state fractions across all time points"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE)

if (!exists("cp_all") || !exists("assign_cells_to_nearest_anchor")) {
  source(here::here("R", "load_and_preprocess.R"))
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
})

ANCHOR_FILE <- here::here("R", "state_anchors.rds")
anchor_bundle <- readRDS(ANCHOR_FILE)
state_anchors <- anchor_bundle$anchors
state_scaling <- anchor_bundle$scaling

OUT_DIR <- here::here("output")
FIG_PREFIX <- file.path(OUT_DIR, "FigureS_timecourse")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

TREATMENTS_6  <- c("Vehicle","LPS","IFNg","IL1beta","IL6","TNFa")
TREATMENTS_F3 <- c("Vehicle","LPS+IFNg")
STATE_LEVELS  <- c("Elongated","Spread","Ramified","Small_Round")
STATE_COLORS  <- c("Elongated"="#FF85C0","Spread"="#00CED1",
                   "Ramified"="#FFD700","Small_Round"="#8B4513")
PAL_6         <- c("Vehicle"="black","LPS"="#AB3A29","IFNg"="#D07F2C",
                   "IL1beta"="#1E7C4A","IL6"="#6F6DA1","TNFa"="#6990A2",
                   "LPS+IFNg"="#7B1E1E")
```

```{r compute-state-fractions-all-timepoints}
# Assign EVERY cell in the pipeline (all stimuli, all time points, IL34+) 
# to its nearest anchor. No downsampling, no UMAP. Outputs:
#   - frac_image: fraction per image
#   - frac_exp:   experiment-level mean of image fractions
#   - frac_summary: mean ± SE across 3 experiments

all_times   <- sort(unique(meta_all$Time_hr))
stimuli_all <- union(TREATMENTS_F3, TREATMENTS_6)

log_cols_actual <- intersect(LOG_COLS, final_features)

sc_timecourse <- cp_all %>%
  inner_join(
    meta_all %>% select(ImageNumber, Experiment, Stimulus, IL34, Time_hr, Well),
    by = "ImageNumber"
  ) %>%
  filter(IL34 == "pos", Stimulus %in% stimuli_all,
         complete.cases(across(all_of(final_features)))) %>%
  mutate(across(all_of(log_cols_actual), log1p)) %>%
  left_join(exp_offset, by = "Experiment", suffix = c("", ".exp")) %>%
  mutate(across(all_of(final_features),
                ~ .x - get(paste0(cur_column(), ".exp")))) %>%
  select(-ends_with(".exp"))

cat(sprintf("Total cells in time course: %d\n", nrow(sc_timecourse)))

sc_timecourse <- sc_timecourse %>% add_morphology_axes(scaling = state_scaling)
state_axes <- c("z_elong","z_size","z_round")
sc_timecourse <- sc_timecourse[complete.cases(sc_timecourse[, state_axes]), ]
sc_timecourse$State <- factor(
  assign_cells_to_nearest_anchor(sc_timecourse, state_anchors),
  levels = STATE_LEVELS)

# Image-level state fractions
frac_image <- sc_timecourse %>%
  count(Time_hr, Experiment, Stimulus, ImageNumber, State) %>%
  complete(nesting(Time_hr, Experiment, Stimulus, ImageNumber),
           State = STATE_LEVELS, fill = list(n = 0)) %>%
  group_by(Time_hr, Experiment, Stimulus, ImageNumber) %>%
  mutate(Fraction = n / sum(n)) %>% ungroup()

# Experiment-level (mean across images)
frac_exp <- frac_image %>%
  group_by(Time_hr, Experiment, Stimulus, State) %>%
  summarise(Fraction = mean(Fraction), .groups = "drop")

# Group-level (mean ± SE across 3 experiments)
frac_summary <- frac_exp %>%
  group_by(Time_hr, Stimulus, State) %>%
  summarise(mean_f = mean(Fraction),
            sd_f   = sd(Fraction),
            n      = n(),
            se_f   = sd_f / sqrt(n),
            .groups = "drop")

write.csv(frac_summary, paste0(FIG_PREFIX, "_summary.csv"), row.names = FALSE)
write.csv(frac_exp,     paste0(FIG_PREFIX, "_per_experiment.csv"), row.names = FALSE)
cat("\nSaved summary CSVs\n")
```

```{r plot-fig3-context, fig.width=10, fig.height=7}
# Figure 3 context: Vehicle vs LPS+IFNg
df_f3 <- frac_summary %>% filter(Stimulus %in% TREATMENTS_F3)

p_f3 <- ggplot(df_f3, aes(Time_hr, mean_f, color = Stimulus, fill = Stimulus)) +
  geom_ribbon(aes(ymin = pmax(0, mean_f - se_f), ymax = mean_f + se_f),
              alpha = 0.2, color = NA) +
  geom_vline(xintercept = c(4, 24), linetype = "dashed",
             color = "grey70", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ State, ncol = 2) +
  scale_color_manual(values = PAL_6,
                     labels = c("Vehicle", expression("LPS+"*IFN*gamma))) +
  scale_fill_manual(values = PAL_6, guide = "none") +
  scale_x_continuous(breaks = all_times) +
  theme_classic(base_size = 11) +
  labs(title = "State fraction time course — Vehicle vs LPS+IFNγ",
       subtitle = "Mean ± SE across 3 experiments; dashed lines = pre-specified 4h, 24h",
       x = "Time (h)", y = "State fraction") +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95"))

p_f3

ggsave(paste0(FIG_PREFIX, "_Fig3context.png"), p_f3 & transparent_theme,
       width = 10, height = 7, dpi = 600, bg = "transparent")
ggsave(paste0(FIG_PREFIX, "_Fig3context.pdf"), p_f3 & transparent_theme,
       width = 10, height = 7, bg = "transparent")
```

```{r plot-fig5-context, fig.width=12, fig.height=7}
# Figure 5 context: 6 conditions
df_f5 <- frac_summary %>% filter(Stimulus %in% TREATMENTS_6)

p_f5 <- ggplot(df_f5, aes(Time_hr, mean_f, color = Stimulus, fill = Stimulus)) +
  geom_ribbon(aes(ymin = pmax(0, mean_f - se_f), ymax = mean_f + se_f),
              alpha = 0.15, color = NA) +
  geom_vline(xintercept = c(4, 24), linetype = "dashed",
             color = "grey70", linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~ State, ncol = 2) +
 scale_color_manual(
  values = PAL_6,
  labels = c(
    "Vehicle" = "Vehicle",
    "LPS"     = "LPS",
    "IFNg"    = expression(IFN*gamma),
    "IL1beta" = expression(IL-1*beta),
    "IL6"     = "IL-6",
    "TNFa"    = expression(TNF*alpha)
  )
) +
  scale_fill_manual(values = PAL_6, guide = "none") +
  scale_x_continuous(breaks = all_times) +
  theme_classic(base_size = 11) +
  labs(title = "State fraction time course — 6 cytokines",
       subtitle = "Mean ± SE across 3 experiments; dashed lines = pre-specified 4h, 24h",
       x = "Time (h)", y = "State fraction") +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95"))

p_f5

ggsave(paste0(FIG_PREFIX, "_Fig5context.png"), p_f5 & transparent_theme,
       width = 12, height = 7, dpi = 600, bg = "transparent")
ggsave(paste0(FIG_PREFIX, "_Fig5context.pdf"), p_f5 & transparent_theme,
       width = 12, height = 7, bg = "transparent")
```

```{r}
```{r plot-figS3-silhouette-diagnostics, fig.width=8, fig.height=3.8}
# Figure S3: k-sensitivity diagnostics for morphological anchor states
# This uses the diagnostic RDS saved by R/define_state_anchors.R

SILHOUETTE_FILE <- here::here("output", "FigureS3_silhouette.rds")

if (!file.exists(SILHOUETTE_FILE)) {
  stop("Silhouette diagnostic file not found: ", SILHOUETTE_FILE,
       "\nRun source(here::here('R', 'define_state_anchors.R')) first.")
}

sil_bundle <- readRDS(SILHOUETTE_FILE)

silhouette_df <- sil_bundle$silhouette
diagnostic_df <- sil_bundle$diagnostic

if (!exists("transparent_theme")) {
  transparent_theme <- theme(
    plot.background   = element_rect(fill = "transparent", colour = NA),
    panel.background  = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)
  )
}

# Panel A: mean silhouette width for k = 2..8
p_sil <- ggplot(silhouette_df, aes(x = k, y = silhouette)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  geom_point(
    data = silhouette_df %>% filter(k == 4),
    size = 3
  ) +
  scale_x_continuous(breaks = 2:8) +
  theme_classic(base_size = 10) +
  labs(
    title = "A",
    x = "Number of clusters (k)",
    y = "Mean silhouette width"
  )

# Panel B: how k = 2 and k = 3 merge the final k = 4 states
diagnostic_long <- diagnostic_df %>%
  pivot_longer(
    cols = c(k2_cluster, k3_cluster),
    names_to = "solution",
    values_to = "cluster"
  ) %>%
  mutate(
    solution = recode(
      solution,
      k2_cluster = "k = 2",
      k3_cluster = "k = 3"
    ),
    State_final = factor(State_final, levels = STATE_LEVELS)
  ) %>%
  count(solution, cluster, State_final) %>%
  group_by(solution, cluster) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()

p_merge <- ggplot(diagnostic_long, aes(x = cluster, y = fraction, fill = State_final)) +
  geom_col(width = 0.8) +
  facet_wrap(~ solution, scales = "free_x") +
  scale_fill_manual(values = STATE_COLORS, drop = FALSE) +
  theme_classic(base_size = 10) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  ) +
  labs(
    title = "B",
    x = "Lower-k cluster",
    y = "Fraction of cells",
    fill = "Final k = 4 state"
  )

p_s3 <- p_sil + p_merge + plot_layout(widths = c(1, 1.6))

p_s3

ggsave(here::here("output", "FigureS3_silhouette_diagnostics.png"),
       p_s3 & transparent_theme,
       width = 8, height = 3.8, dpi = 600, bg = "transparent")

ggsave(here::here("output", "FigureS3_silhouette_diagnostics.pdf"),
       p_s3 & transparent_theme,
       width = 8, height = 3.8, bg = "transparent")

write.csv(silhouette_df,
          here::here("output", "FigureS3_silhouette_scores.csv"),
          row.names = FALSE)

write.csv(diagnostic_long,
          here::here("output", "FigureS3_lower_k_diagnostics.csv"),
          row.names = FALSE)
```


